\chapter{Conclusions and future work}

%I briefly overview the architecture of the VIATRA-CEP framework.
The architecture of the VIATRA-CEP was briefly overviewed.
%I introduce multiple automata and regular expression formalisms, and specify transformations from the regular expressions to the automata.
Multiple automata and regular expression formalisms were overviewed, and some specifications were introduced from these regular expressions to the corresponding automata formalisms. 
%And lastly, I go through the implementation of these formalisms and transformations, and show the transformations from the language of VIATRA-CEP to one of the previously mentioned formalism.
After all these introductions, the implementation of these formalisms and transformations were overviewed, a transformation was shown from the language of VIATRA-CEP to the Parametric Timed Regular Expressions.



\section{Future Work}

There are many ways to optimize these automata, and the research about this topic is considered future work. 
There could be algorithms which can remove $\varepsilon$-edges from an automaton, and the management of the tokens could be also improved via merging tokens by some logic.

The formalism could be extended, for example towards the State Charts instead of the plain automata, as it could provide hierarchy. 

\paragraph*{Acknowledgments}
\phantomsection
\addcontentsline{toc}{chapter}{Acknowledgments}


First and foremost, I would like to thank my supervisor, \advisori, as without his guidance in the last five years I could not create such a work as this thesis.

I would also like to thank my two other advisors \advisorii, and \advisoriii, without their help in formalization and algorithmization would have been way more cumbersome.

I would like to thank Oszk치r Semer치th as without his help the measurements would have been a lot harder.

And last, but not least, I would like to thank Krist칩f Marussy, and B치lint Hegyi for helping me with L\textsc{a}T\textsc{e}X and Ti\emph{k}Z related issues, to help me create such a beautiful document.

